Exoplanet population inference \& the rate of Earth analogs from noisy, incomplete catalogs

### 2014-06-01 document/ms.tex
style note
small intro tweaks

%
%  RULES OF THE GAME
%
%  * 80 characters
%  * line breaks at the ends of sentences
%  * eqnarrys ONLY
%  * that is all.
%

@@

NASA's \kepler\ mission has enabled the discovery of thousands of exoplanet
candidates (\citealt{kepler-catalog}).
While many of these candidates have not been confirmed as bona fide planets,
there is evidence that the false positive rate is low (\citealt{morton,
fressin-fp}), enabling conclusions about the population of planets based on
the catalog of candidates.
Many of these planets orbit Sun-like stars (\citealt{petigura}), where the
definition of Sun-like is given in terms of the star's temperature and surface
gravity.
Given these catalogs, it is interesting to ask what we can say about the
occurrence rate of exoplanets as a function of their physical parameters
(period, radius, \etc).
Observational constraints on the occurrence rate function can inform theories
of planet formation (\dfmtodo{CITE}) and place probabilistic bounds on the
rate of Earth analogs\footnote{We define, in detail, what we mean by Earth
analog below but, in short, we mean Earth-sized planets on year-long orbits
around Sun-like stars.}.

@@

In every previous study of the occurrence rate function, the authors have
assumed that the measurement uncertainties are negligible.
This assumption is not justified because these uncertainties---especially on
measurements of the stellar parameters---can be large compared to the scales
of interest.
In this \paper, we develop a flexible framework for probabilistic inference of
exoplanet occurrence rates that can be applied to incomplete catalogs with
\emph{non-negligible observational uncertainties}.
We generalize the inference method introduced by \citet{hogge} to account for
survey detection efficiencies.
We run tests on simulated datasets---comparing results with the standard
techniques that neglect observational uncertainties---and apply our method to
a real catalog of small planets transiting Sun-like stars
(\citealt{petigura}).

For the purposes of this \paper, we make some strong assumptions---although we
argue that they are weaker than the implicit assumptions in previous
studies.
None of these assumptions are necessary for the applicability of our method
but they do simplify the procedure.
Specifically, we assume that
\begin{itemize}

@@

\end{itemize}
The first assumption---conditional independence of the candidates---is
reasonable since the dataset that we consider explicitly includes only single
transiting systems (\citealt{petigura}).
When considering other datasets it will be interesting and important to take
multiple transiting systems into account (\citealt{lissauer, tremaine, fang})
and this would be possible using a conditional distributions (as developed by
\citealt{tremaine}).
The second assumption---neglecting false positives---is also strong and only
weakly justified by estimates of low false positive rates in the \kepler\ data
(\citealt{fressin-fp, morton}).
For this \paper, we will neglect this issue and only comment on the effects
but the prior distributions published by \citet{fressin-fp} would be directly
applicable in our method and this would be a very interesting generalization.

We must emphasize one very important consequence of our assumptions.
We assume that the catalog of exoplanet candidates is only missing planets
with probabilities given by the empirical detection efficiency but this
doesn't take into account the fact that the catalog doesn't include multiple
transiting systems.
A large fraction of the transiting planets discovered by the \kepler\ transit
search pipeline are members of multiple transiting systems (see
\citealt{lissauer}, for example).
Since \citet{petigura} only detected at most one planet per system, their
catalog is actually a list of planet candidates \emph{without a more
detectable companion}.
The global effects of this selection is not trivial and an in-depth discussion
is beyond the scope of this \paper\ but all of the results should be
interpreted with this caveat in mind.

Conditioned on our assumptions and the choices made in planet detection,
vetting and characterization pipeline (\citealt{petigura-a, petigura}), we
constrain the rate of small exoplanets orbiting Sun-like stars.
As part of this analysis we also place probabilistic constraints on the rate
of Earth analogs \gammaearth\ which we define as \emph{the number of planets
per star per natural logarithmic bin in period and radius, evaluated at the
period and radius of Earth}
\begin{eqnarray}
\gammaearth &=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
Since no Earth analogs have been detected, this constraint requires an
extrapolation in both period and radius.
\citet{petigura} performed this extrapolation by assuming that the period
distribution of planets in a small bin in radius is flat, obtaining
$\gammaearth \approx 0.12$.
We relax this assumption and extrapolate by only assuming that the occurrence
rate is a smooth function of period and radius; we find a much lower value for
\gammaearth.
We enforce the smoothness constraint by applying a flexible Gaussian process
regularization to the bin heights.

In the next section, we introduce the concept of hierarchical inference, how
it applies to the problem at hand, and we derive the importance sampling
approximation to the marginalized likelihood (this is a generalization of the
method from \citealt{hogge}).
Then, in \sect{model}, we describe how to include detection efficiencies in
this model and derive a probabilistic motivation for the standard
inverse-detection-efficiency weighting procedure.
After testing our method on synthetic catalogs in \sect{valid}, we use the
catalog of planet candidates and the empirically determined detection
efficiency from \citet{petigura} to measure the occurrence rate of Earth
analogs.

>>>>

%
%  RULES OF THE GAME
%
%  * 80 characters
%  * line breaks at the ends of sentences
%  * Sun Solar Earth all capitalized when referring to our peeps, even as in "extra-Solar"
%  * eqnarrys ONLY
%  * that is all.
%

@@

NASA's \kepler\ mission has enabled the discovery of thousands of exoplanet
candidates (\citealt{kepler-catalog}).
While many of these candidates have not been confirmed as bona fide planets,
there is evidence that the false positive rate is low (\citealt{morton,
fressin-fp}), enabling conclusions about the population of planets based on
the catalog of candidates.
Many of these planets orbit Sun-like stars (\citealt{petigura}), where the
definition of Sun-like is given in terms of the star's temperature and surface
gravity.
Given these catalogs, it is interesting to ask what we can say about the
occurrence rate of exoplanets as a function of their physical parameters
(period, radius, \etc).
Observational constraints on the population can inform theories
of planet formation (\dfmtodo{CITE}) and place probabilistic bounds on the
rate of Earth analogs\footnote{We define, in detail, what we mean by ``Earth
analog'' below but, roughly, we mean Earth-sized planets on year-long orbits
around Sun-like stars.}.

@@

In every previous study of the occurrence rate function, the authors have
assumed that the measurement uncertainties are negligible.
This assumption is not justified because these uncertainties---especially on
measurements (like exoplanet radius) that depend on the stellar
parameters---can be large compared to the scales of interest.
In this \paper, we develop a flexible framework for probabilistic inference of
exoplanet occurrence rates that can be applied to incomplete catalogs with
\emph{non-negligible observational uncertainties}.
We generalize the inference method introduced by \citet{hogge} to account for
survey detection efficiencies.
We run tests on simulated datasets---comparing results with the standard
techniques that neglect observational uncertainties---and apply our method to
a real catalog of small planets transiting Sun-like stars
(\citealt{petigura}).

For the purposes of this \paper, we make some strong assumptions, although we
argue that they are weaker than the implicit assumptions in previous
studies.
None of these assumptions is necessary for the validity of our general method
but they do simplify the specific procedures we employ.
Specifically, we assume that
\begin{itemize}

@@

\end{itemize}
The first assumption---conditional independence of the candidates---is
reasonable since the dataset that we consider explicitly includes only single
transiting systems (\citealt{petigura}).
This is a limitation of the input catalog.
When considering other datasets it will be interesting and important to take
multiple transiting systems into account (\citealt{lissauer, tremaine, fang})
and this would be possible using a conditional distributions (as developed by
\citealt{tremaine}).
The second assumption---neglecting false positives---is also strong and only
weakly justified by estimates of low false positive rates in the \kepler\ data
(\citealt{fressin-fp, morton}).
For this \paper, we will neglect this issue and only comment on the effects
but the prior distributions published by \citet{fressin-fp} would be directly
applicable in our method and this would be a very interesting generalization.

We must emphasize one very important consequence of our assumptions.
We assume that the catalog of exoplanet candidates is only missing planets
with probabilities given by the empirical detection efficiency.
In detail this must be false; the detection efficiency we use
doesn't take into account the fact that the catalog doesn't include multiple
transiting systems.
A large fraction of the transiting planets discovered by the \kepler\ transit
search pipeline are members of multiple transiting systems (see
\citealt{lissauer}, for example).
Since \citet{petigura} only detected at most one planet per system, their
catalog is actually a list of planet candidates \emph{without a more
detectable companion}.
The global effects of this selection is not trivial and an in-depth discussion
is beyond the scope of this \paper\ but all of the results should be
interpreted with this caveat in mind.

Conditioned on our assumptions and the choices made in planet detection,
vetting and characterization pipeline (\citealt{petigura-a, petigura}), we
constrain the rate of small exoplanets orbiting Sun-like stars.
As part of this analysis we also place probabilistic constraints on the rate
of Earth analogs \gammaearth\ which we define as \emph{the number of planets
per star per natural logarithmic bin in period and radius, evaluated at the
period and radius of Earth} % Prof Tremaine, you owe us a beer.
\begin{eqnarray}
\gammaearth &=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
Since no Earth analogs have been detected, this constraint requires an
extrapolation in both period and radius.
\citet{petigura} performed this extrapolation by assuming that the period
distribution of planets in a small bin in radius is flat, obtaining
$\gammaearth \approx 0.12$.
We relax this assumption and extrapolate only by assuming that the occurrence
rate is a smooth function of period and radius; we find a much lower value for
\gammaearth.
We enforce the smoothness constraint by applying a flexible Gaussian process
regularization to the bin heights.

In the next \sectionname, we introduce the concept of hierarchical inference, how
it applies to the problem at hand, and we derive the importance sampling
approximation to the marginalized likelihood (this is a generalization of the
method from \citealt{hogge}).
Then, in \sect{model}, we describe how to include detection efficiencies in
this model and derive a probabilistic motivation for the standard
inverse-detection-efficiency weighting procedure.
After testing our method on synthetic catalogs in \sect{valid}, we use the
catalog of planet candidates and the empirically determined detection
efficiency from \citet{petigura} to measure the occurrence rate of Earth
analogs.
